---
title: "Analysis of Dental Magic's Human Resources Dataset"
author: "Jennie Eardley"
date: "08/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
library(tidyverse)
library(janitor)
library(plotly)
library(gt)
library(tsibble)
library(anytime)

hr <- read_csv("data/HRDataset_v13.csv") %>% 
  clean_names()
hr <- data.frame(lapply(hr, function(x){
  if (is.character(x)) return(tolower(x))
  else return(x)
}))
hr <- hr %>% 
  mutate_if(is.factor, as.character)
hr <- hr %>% 
  rename(c("date_of_hire" = "dateof_hire", "date_of_termination" = "dateof_termination", "days_late_last_30" = "days_late_last30"))
hr <- hr %>% 
  filter(!is.na(emp_id))
```

**Data Source**

This is a dataset based around a fictional company called Dental Magic, created to teach HR analytics. The dataset is from Kaggle and is available here: https://www.kaggle.com/rhuebner/human-resources-data-set

**Recruitment**
 

```{r echo = FALSE}
hr <- hr %>% 
  mutate(date_of_hire = anydate(date_of_hire))

hr_tsibble <- as_tsibble(hr, key = emp_id, index = date_of_hire)

indexed_tsibble <- hr_tsibble %>% 
  index_by(hire_month = ~ yearmonth(.)) %>% 
  summarise(monthly_count = n()) 

ggplot(indexed_tsibble, aes(x = hire_month, y = monthly_count)) +
  geom_line(group = 1, colour = "dark blue") +
  labs(x = "Year",
       y = "Count",
       title = "Number of New Staff Recruited by Time") +
    theme(plot.title = element_text(hjust = 0.5))
```


```{r echo = FALSE}
hr %>% 
  mutate(recruitment_source = fct_rev(recruitment_source)) %>% 
  filter(!is.na(recruitment_source)) %>%
  ggplot() +
  geom_bar(aes(x = recruitment_source)) +
  coord_flip() +
  labs(x = "Count",
       y = "Recruitment Source",
       title = "Recruitment Methods") +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
hr %>% 
  filter(!is.na(emp_satisfaction)) %>% 
  ggplot() +
  geom_point(aes(x = emp_satisfaction, y = pay_rate))
```


**Diversity**

```{r echo = FALSE}
 hr %>% 
  filter(!is.na(race_desc)) %>% 
  mutate(race_desc = fct_rev(race_desc)) %>% 
  ggplot() +
  geom_bar(aes(x = race_desc), fill = "dark blue") + 
  coord_flip() +
  labs(x = "Racial Description",
       y = "Count", 
       title = "Racial Breakdown Across the Company") +
    theme(plot.title = element_text(hjust = 0.5))
```
```{r echo = FALSE}
hr %>% 
  mutate(race_desc = fct_rev(race_desc)) %>% 
  filter(from_diversity_job_fair_id == 1) %>% 
  ggplot() + 
  geom_bar(aes(x = race_desc), fill = "dark blue") +
  coord_flip() +
  labs(x = "Racial Description", 
      y = "Count",
      title = "Racial Breakdown from Diversity Job Fair Recruitment")
```
**Employee Satisfaction Levels**

**Gender Pay Gap**

On average, men earn $33.70 compared to $29.47 for women. 
```{r echo = FALSE}
hr %>% 
  filter(sex == "m") %>% 
  summarise(n = mean(pay_rate)) %>% 
  round(digits = 2)
```
```{r echo = FALSE}
hr %>% 
  filter(sex == "f") %>% 
  summarise(n = mean(pay_rate)) %>% 
    round(digits = 2)
```
```{r}
male_rate <- hr %>% 
  filter(sex == "m") %>% 
  group_by(position) %>% 
  summarise(average_male_rate = mean(pay_rate)) 

female_rate <- hr %>% 
  filter(sex == "f") %>% 
  group_by(position) %>% 
  summarise(average_female_rate = mean(pay_rate)) 

average_rate <- inner_join(male_rate, female_rate) %>% 
  mutate(difference = average_male_rate - average_female_rate) %>% 
  arrange(desc(difference))
```
```{r}

```



**Staff Turnover**

There are a total of 310 staff records, with 196 currently employed and 11 due to onboard. The rest have all left, the breakdown for their reasons of leaving are as follows:

```{r echo = FALSE}
hr %>% 
  mutate(term_reason = fct_rev(term_reason)) %>% 
  filter(!is.na(term_reason)) %>% 
  filter(term_reason != "n/a - still employed") %>% 
  filter(term_reason != "n/a - has not started yet") %>% 
  ggplot() +
  geom_bar(aes(x = term_reason), fill = "dark blue") +
  coord_flip() +
  labs(x = "Reason",
       y = "Count",
       title = "Reason for Leaving") +
    theme(plot.title = element_text(hjust = 0.5))

```
**Recommendations for Future Data Collection**

