---
title: <center> Analysis of Dental Magic's Human Resources Data <center>
author: <center> Jennie Eardley <center>
date: <center> 08/10/2020 <center>
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

```{r include = FALSE}
library(tidyverse)
library(janitor)
library(plotly)
library(gt)
library(tsibble)
library(anytime)
library(lubridate)

hr <- read_csv("data/HRDataset_v13.csv") %>% 
  clean_names()

# turns data into lower case 
hr <- data.frame(lapply(hr, function(x){
  if (is.character(x)) return(tolower(x))
  else return(x)
}))

# switches data back to characters
hr <- hr %>% 
  mutate_if(is.factor, as.character)

# renames columns for consistency
hr <- hr %>% 
  rename(c("date_of_hire" = "dateof_hire", "date_of_termination" = "dateof_termination", "days_late_last_30" = "days_late_last30"))

# removes empty rows
hr <- hr %>% 
  filter(!is.na(emp_id))
```


**Data Source**

This is a dataset based around a fictional company called Dental Magic, created to teach HR analytics. The dataset is from Kaggle and is available here: https://www.kaggle.com/rhuebner/human-resources-data-set

**Recruitment**
 
 The data covers the time period of 2006 to 2017. This graph shows the number of new hires by time. The month with the highest number of new recruits was in September of 2011 with 15 new recruits. 
 
```{r fig.align='center', echo = FALSE}
# switching the dates to a consistent format
hr <- hr %>% 
  mutate(date_of_hire = anydate(date_of_hire))

# converting the data to a tsibble to enable the creation of a timeseries 
hr_tsibble <- as_tsibble(hr, key = emp_id, index = date_of_hire)

suppressWarnings(indexed_tsibble <- hr_tsibble %>% 
  index_by(hire_month = ~ yearmonth(.)) %>% 
  summarise(monthly_count = n(), .groups = "drop"))

ggplot(indexed_tsibble, aes(x = hire_month, y = monthly_count)) +
  geom_line(group = 1, colour = "dark green") +
  labs(x = "Year",
       y = "Number of New Employees",
       title = "Timing of New Staff Recruitment") +
    theme(plot.title = element_text(hjust = 0.5))
```

It is possible to group the hires together by month. January is clearly the most popular month for hiring. This is perhaps unsurprising, compensating for the low hire rate in December. The leavers' data also shows that January is the most common month for employees to leave so the January spike in recruitment could simply be replacing those that have left. This would indicate a quick recruitment turnaround process. 

```{r fig.align='center', echo = FALSE}
# providing a monthly breakdown
indexed_monthly_tsibble <- hr_tsibble %>% 
  index_by(monthly_hire = ~ month(., label = TRUE)) %>% 
  summarise(seasonal_count = n(), .groups = "drop")

ggplot(indexed_monthly_tsibble, aes(x = monthly_hire, y = seasonal_count)) +
  geom_col(fill = "dark green") +
  labs(x = "Month",
       y = "Number of New Hires",
       title = "Recruitment by Month") +
    theme(plot.title = element_text(hjust = 0.5))
```

The most common recruitment route is via diversity job fairs, with search engines closely following. Other targeted methods of recruitment seem to be less successful, particularly on campus recruiting, web banner ads and information sessions. It may be worth considering funds spent on these activities and their content. 

```{r fig.align='center', echo = FALSE}
hr %>% 
  mutate(recruitment_source = fct_rev(recruitment_source)) %>% 
  filter(!is.na(recruitment_source)) %>%
  ggplot() +
  geom_bar(aes(x = recruitment_source), fill = "dark green") +
  coord_flip() +
  labs(x = "Count",
       y = "Recruitment Source",
       title = "Recruitment Methods") +
  theme(plot.title = element_text(hjust = 0.5))
```

**Diversity in Recruitment**

As well as introducing a large number of successful candidates to the company, the data indicates that participation in the Diversity Job Fair is successful in attracting more racially diverse candidates. 

```{r echo = FALSE, figures-side, fig.show = "hold", out.width = "50%"}
# reversing the factors to ensure alphabetisation when the graph is flipped
 hr %>% 
  filter(!is.na(race_desc)) %>% 
  mutate(race_desc = fct_rev(race_desc)) %>% 
  ggplot() +
  geom_bar(aes(x = race_desc), fill = "dark green") + 
  coord_flip() +
  labs(x = "Racial Description",
       y = "Number of Employees", 
       title = "Racial Breakdown Across the Company") +
    theme(plot.title = element_text(hjust = 0.5))

hr %>% 
  mutate(race_desc = fct_rev(race_desc)) %>% 
  filter(from_diversity_job_fair_id == 1) %>% 
  ggplot() + 
  geom_bar(aes(x = race_desc), fill = "dark green") +
  coord_flip() +
  labs(x = "Racial Description", 
      y = "Number of Employees",
      title = "Racial Breakdown from Diversity Job Fair Recruitment") +
  theme(plot.title = element_text(hjust = 0.5))
```

**Employee Satisfaction** 

Employee satisfaction is measured on a scale of 1 to 5, with 5 indicating the highest satisfaction. Across the dataset, the employee satisfaction median score is 4 with a mode score of 3.

The data gives numerous factors which could have an impact on employee satisfaction. This report considers the impact of different factors in influencing employee satisfaction. Firstly, the impact of pay on employee satisfaction. The plot does not suggest that there is much of a correlation between employee satisfaction and their payrate as the data is very spread out. 

```{r fig.align='center', echo = FALSE}
hr %>% 
  filter(!is.na(emp_satisfaction)) %>% 
  ggplot() +
  geom_point(aes(x = emp_satisfaction, y = pay_rate), col = "dark green") +
  labs(x = "Employee Satisfaction Score",
       y = "Pay Rate",
       title = "Employee Satisfaction by Pay") +
  theme(plot.title = element_text(hjust = 0.5))

```

<br>

```{r fig.align='center', echo = FALSE}
satisfaction_by_position <- hr %>% 
  group_by(position) %>% 
  summarise(average_score = round(mean(emp_satisfaction, .groups = "drop"), digits = 2))

position_count <- hr %>% 
  count(position)

position_satisfaction <- full_join(satisfaction_by_position, position_count) %>% 
  rename("number_of_employees" = "n") %>% 
  filter(number_of_employees > 1) %>% 
  arrange(desc(average_score))

position_satisfaction <- position_satisfaction %>% 
  rename("Number of Employees" = "number_of_employees", "Position" = "position", "Average Score" = "average_score")
```

```{r echo = FALSE}
total_employees <- hr %>% nrow()

fully_meets <- hr %>% 
  filter(performance_score == "fully meets") %>% 
  nrow()

exceeds <- hr %>% 
  filter(performance_score == "exceeds") %>% 
  nrow()

pip <- hr %>% 
  filter(performance_score == "pip") %>% 
  nrow()

needs_improvement <- hr %>% 
  filter(performance_score == "needs improvement") %>% 
  nrow()

fully_meets_percentage <- round(fully_meets/total_employees * 100)
exceeds_percentage <- round(exceeds/total_employees * 100)
pip_percentage <- round(pip/total_employees * 100)
needs_improvement_percentage <- round(needs_improvement/total_employees * 100)

```

This table shows average satisfaction levels split by employee position. It also includes the number of employees that work in that position. It excludes positions that only have one employee. 

<br>
```{r echo = FALSE}
# creates a formatted table 
gt(position_satisfaction) %>% 
  tab_header("Average Satisfaction Scores Grouped by Position", subtitle = "Where Position Has More Than One Employee")
```

<br>

**Performance**

This graph shows the overall picture of performance within the company. While it does include those who have left, it excludes those who are listed but have not yet started. They seem to have erroneously been assigned a performance category and removing them prevents inaccuracies in performance score reporting. 

The vast majority of employees' performance is rated as "fully meets" (78%). 12% of employees have a score of exceeds and 6% and 4% have scores of needs improvement and pip respectively. 

```{r fig.align = 'center', echo = FALSE}
# removing employees with a future start date to ensure that data is only from people who have actually worked for the company
hr %>% 
  filter(employment_status != "future start") %>% 
  ggplot() +
  geom_bar(aes(x = performance_score, fill = employment_status)) +
  labs(x = "Score",
       y = "Count",
       title = "Performance Scores by Employment Status",
       fill = "Employment Status") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_fill_brewer(palette = "Greens")
```

It is notable that while employees with the highest performance level of "exceeds" is the only category where the mode is 5, they do not seem to be significantly more satisfied than those with other performance levels. It is unsuprising that employees with the "needs improvement" and "pip" performance levels have disproportionately incidence of a satisfaction level of 2 and two people have reported a satisfaction level of 1.

```{r fig.align = 'center', echo = FALSE}
hr %>% 
  ggplot() +
  geom_bar(aes(x = emp_satisfaction), fill = "dark green") +
  facet_wrap(~ performance_score) +
  labs(x = "Employee Satisfaction",
       y = "Count",
       title = "Employee Satisfaction Levels Separated by Performance Levels")
```

**Gender Pay Gap**

On average, men earn $33.70 compared to $29.47 for women. On average, men earn more than women in nine different positions. The position with the highest discrepancy is IT support workers where men earn a mean of $38.66 compared to $32.18 for women. On average, women earn more than men in six different positions with the largest discrepancy being for software engineers where women earn an average of 52.33 compared to the 48.56 earned by men. 

```{r include = FALSE}
hr %>% 
  filter(sex == "m") %>% 
  summarise(n = mean(pay_rate)) %>% 
  round(digits = 2)
```
```{r include = FALSE}
hr %>% 
  filter(sex == "f") %>% 
  summarise(n = mean(pay_rate)) %>% 
    round(digits = 2)
```
```{r include = FALSE}
male_rate <- hr %>% 
  filter(sex == "m") %>% 
  group_by(position) %>% 
  summarise(average_male_rate = mean(pay_rate)) 

female_rate <- hr %>% 
  filter(sex == "f") %>% 
  group_by(position) %>% 
  summarise(average_female_rate = mean(pay_rate)) 

average_rate <- inner_join(male_rate, female_rate) %>% 
  mutate(difference = average_male_rate - average_female_rate) %>% 
  arrange(desc(difference))
```

**Turnover**

It also highlights two positions where the sole employee has not been replaced. 

```{r echo = FALSE}
have_left <- hr %>% 
   filter(!is.na(term_reason)) %>% 
  filter(term_reason != "n/a - still employed") %>% 
  filter(term_reason != "n/a - has not started yet")
have_left

total_positions <- hr %>% 
  count(position) 

positions_left <- have_left %>% 
count(position) 

right_join(total_positions, positions_left, by = c("position" = "position")) %>%
  rename("Total" = "n.x", "Have_Left" = "n.y", "Position" = "position") %>% 
  mutate(Turnover = round(Have_Left/Total * 100, digits = 2)) %>% 
  arrange(desc(Turnover)) %>% 
  gt()
```


**Employees Who Have Left**

There are a total of 310 staff records, with 196 currently employed and 11 due to onboard. The rest have all left, the breakdown for their reasons of leaving are as follows:

```{r fig.align = 'center', echo = FALSE}
hr %>% 
  mutate(term_reason = fct_rev(term_reason)) %>% 
  filter(!is.na(term_reason)) %>% 
  filter(term_reason != "n/a - still employed") %>% 
  filter(term_reason != "n/a - has not started yet") %>% 
  ggplot() +
  geom_bar(aes(x = term_reason), fill = "dark green") +
  coord_flip() +
  labs(x = "Reason",
       y = "Count",
       title = "Reason for Leaving") +
    theme(plot.title = element_text(hjust = 0.5))

```

As with timing of recruiting, we can also evaluate the timing of staff leaving. 

```{r fig.align = 'center', echo = FALSE}
left_hr <- hr %>% 
   filter(!is.na(date_of_termination)) %>% 
  mutate(date_of_termination = mdy(date_of_termination)) 

left_tsibble <- as_tsibble(left_hr, key = emp_id, index = date_of_termination) 
 
left_indexed_tsibble <- left_tsibble %>% 
  index_by(leaving_month = ~ yearmonth(.)) %>% 
  summarise(monthly_count = n(), .groups = "drop") 

ggplot(left_indexed_tsibble, aes(x = leaving_month, y = monthly_count)) +
  geom_line(group = 1, colour = "dark green") +
  labs(x = "Year",
       y = "Number of Employees",
       title = "Timing of Staff Leaving") +
    theme(plot.title = element_text(hjust = 0.5))
```

This can also be broken down to identify months with a particularly high rate of leavers. As mentioned at the beginning of the report, the most popular month to leave is during January. 

```{r fig.align = 'center', echo = FALSE}
# calculating staff leaving by month
indexed_left_monthly_tsibble <- hr_tsibble %>% 
  index_by(monthly_loss = ~ month(., label = TRUE)) %>% 
  summarise(seasonal_loss = n(), .groups = "drop")

ggplot(indexed_left_monthly_tsibble, aes(x = monthly_loss, y = seasonal_loss)) +
  geom_col(fill = "dark green") +
  labs(x = "Month",
       y = "Number of Staff that Left",
       title = "Staff Leaving Numbers by Month") +
    theme(plot.title = element_text(hjust = 0.5))
```
